A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-10-11 paper/paper.discussion.tex
Update paper.discussion.tex
SM reviewed discussion
Cleanup latex comments

In this paper, we present a simple, yet flexible, integer linear programming
formulation to generate schedules for clinical departments at hospitals.
The challenge in applying ILP to the task of scheduling clinicians lies in the
computational complexity of finding an optimal solution. As the size of the
% JK: Is this a challenge of applying ILP? Or of the problem in general?
%     I got the impression that ILP was a nice way to overcome this problem.
scheduling problem grows, due to a larger roster of clinicians or more
complicated constraints, the time it takes to generate an optimal schedule may
grow exponentially in the worst case.
%The challenge in applying such an approach to this task lies in the fact that
%ILP is an NP-hard problem. %specify what 'such an approach' and 'this task'
%are... the first part of the sentence was a bit hard to follow. Define what you
%mean by NP-hard problem and cite --> i.e. NP-hard = the following sentence
%about how time to optimal solution grows exponentially? if yes, then instead of
%'As such' - say 'That is,...or "An NP-hard problem means that..."
% JK: How come only "may"? Under what conditions?
Many previous approaches to creating schedules in similar scenarios have avoided
this problem by using heuristics to find an approximately optimal solution 
in a shorter amount of time~\cite{burke_state_2004}.
%As a result, the majority of approaches taken to create schedules for similar
%scenarios tend to use heuristics in order to find an approximately optimal
%solution in a shorter time~\cite{burke_state_2004}. %cite/reference
%statement (i.e. the 'vast majority'...) I can't remember if the introduction
%details the heuristics --> but if yes, then great. if no, then include that
%review of the prior literature here in discussion. I think as you said - the
%journals we are submitting to seem to have more of those details in the
%introduction.

%double-check all abbreviations have been defined. I would prefer limiting
%abbreviations to no more than 3 at most if possible. We are ok for word count
%in this paper :).

We presented a formulation that includes both hard constraints to ensure the
schedule satisfies hospital and logistics requirements, and a multi-goal
objective function to satisfy soft constraints (work preferences of clinicians).
%Our formulation includes hard constraints to ensure the schedule satisfies
%hospital and logistics requirements. It also aims to satisfy the work
%preferences of clinicians in a clinical department by optimizing a multi-goal
%objective function. 
Although we restricted our application of the formulation to a set of
constraints for the particular needs of the case study (St.\ Michael's Hospital
Division of Infectious Diseases), our formulation can be adapted to various
clinical departments at different hospitals. The flexibility of our ILP allows
changing the number of services provided in a division, the length of a work
block, clinicians' preference for block to weekend adjacency as well as
clinicians' requests for time off.
% JK: Nice!

When comparing the optimal schedule generated by our tool to the
manually-created schedules at St.\ Michael's Hospital, we found that the ILP
formulation was always able to find an optimal schedule satisfying all required
hard constraints, unlike the manual schedule, which often did not satisfy all
constraints. Moreover, due to the multi-goal objective function in
the ILP, the algorithm was able to fulfill the majority of clinician
preferences and requests, more so than the manually-created schedule. These
observations reinforce the benefits of automated tools when generating schedules
in hospital departments to balance the workload of clinicians and improve the
service provided to patients. The use of automated tools alleviates the time
spent on designing the schedule by hand, and provides clinical departments with
a more fair distribution of work that helps improve the overall satisfaction of
both employees and patients~\cite{silvestro_evaluation_2000}.  % how the
%finding compares to the wider literature when automated compared to manual?
%What does it mean re: "so what" - human error in generating tools, etc. or
%challenges in heuristics used manually by people? include citations this last
%sentence in particular - what dos it reinforce exactly - what are the benefits?

In our simulations, we also found that increasing the number of requests per
clinician did not affect the runtime of the algorithm, highlighting the
flexibility of the tool to incorporate clinician preferences. Further, we saw
that the algorithm can accommodate an increase in time-horizon up to four years
with little impact on runtime, suggesting the algorithm can be used generate
schedules far in advance. One key limitation we identified was the sensitivity
of the runtime to larger numbers of services offered by a single department. 
One solution to mitigate the runtime issues created by a
larger number of services would be removing the constraint that prevents
assignment of consecutive blocks, followed by manual readjustment from the
generated schedule. %what do we mean by certain constraints? specify the other
%constraints?
Overall, our sensitivity analyses using simulated data provided reassurance that
the ILP formulation can be applied to schedule clinicians across real-world
variability between clinical departments. %SM - will look for reference and we
%can ask Kevin Gough too.
Next steps include expanding the generalizability of the tool beyond smaller
clinical departments to larger departments within and outside of health-care --
especially those that provide multiple services in parallel for patients and
other clients. As well, additional work can be done to incorporate other
clinician preferences, such as the ability to request time-on slots or preference
for time of year.

>>>>

In this paper, we present a simple, yet flexible, integer linear programming
formulation to generate schedules for clinical departments at hospitals.
The challenge in applying ILP to the task of scheduling clinicians lies in the
computational complexity of finding an optimal solution. As the size of the
scheduling problem grows, due to a larger roster of clinicians or more
complicated constraints, the time it takes to generate an optimal schedule may
grow exponentially in the worst case.
Many previous approaches to creating schedules in similar scenarios have avoided
this problem by using heuristics to find an approximately optimal solution 
in a shorter amount of time~\cite{burke_state_2004}.

We presented a formulation that includes both hard constraints to ensure the
schedule satisfies hospital and logistics requirements, and a multi-goal
objective function to satisfy soft constraints (work preferences of clinicians).
Although we restricted our application of the formulation to a set of
constraints for the particular needs of the case study (St.\ Michael's Hospital
Division of Infectious Diseases), our formulation can be adapted to various
clinical departments at different hospitals. The flexibility of our ILP allows
changing the number of services provided in a division, the length of a work
block, clinicians' preference for block to weekend adjacency as well as
clinicians' requests for time off.

When comparing the optimal schedule generated by our tool to the
manually-created schedules at St.\ Michael's Hospital, we found that the ILP
formulation was always able to find an optimal schedule satisfying all required
hard constraints, unlike the manual schedule, which often did not satisfy all
constraints. Moreover, due to the multi-goal objective function in
the ILP, the algorithm was able to fulfill the majority of clinician
preferences and requests, more so than the manually-created schedule. These
observations reinforce the benefits of automated tools when generating schedules
in hospital departments to balance the workload of clinicians and improve the
service provided to patients. The use of automated tools alleviates the time
spent on designing the schedule by hand, and provides clinical departments with
a more fair distribution of work that helps improve the overall satisfaction of
both employees and patients~\cite{silvestro_evaluation_2000}.

In our simulations, we also found that increasing the number of requests per
clinician did not affect the runtime of the algorithm, highlighting the
flexibility of the tool to incorporate clinician preferences. Further, we saw
that the algorithm can accommodate an increase in time-horizon up to four years
with little impact on runtime, suggesting the algorithm can be used generate
schedules far in advance. A key limitation we identified was the sensitivity
of the runtime to larger numbers of services offered by a single department. 
One solution to mitigate the runtime issues created by a
larger number of services would be removing the constraint that prevents
assignment of consecutive blocks, followed by manual readjustment from the
generated schedule.
Overall, our sensitivity analyses using simulated data provided reassurance that
the ILP formulation can be applied to schedule clinicians across real-world
variability between clinical departments.
Next steps include expanding the generalizability of the tool beyond smaller
clinical departments to larger departments within and outside of health-care --
especially those that provide multiple services in parallel for patients and
other clients. As well, additional work can be done to incorporate other
clinician preferences, such as the ability to request time-on slots or preference
for time of year.

### 2019-10-11 paper/paper.experiments.tex
Cleanup latex comments

% JK: I think give an overview here of the experiments here
%     and what you're going to show in this section (and why)
In this section we compare the schedules created by solving the ILP formulation
given in Section~\ref{sec:methods} to schedules that were manually generated,
in terms of adherence to the hard and soft constraints outlined in Section~\ref{sec:problem}.
We also want to analyze the efficiency of the ILP approach in generating schedules
by its runtime on a variety of instances that may be found in the real-world.
Our goal is to show that the schedules provided by solving the ILP successfully
enforce all hard constraints and improve satisfiability of soft constraints, 
as well as to demonstrate that our ILP formulation can be used for a wide range
of configurations.

\subsection{Implementation}
We developed a Python software package with a user interface that implements the above
linear program and allows configuration of clinicians, to be
% JK: what is this ref? Do you want to mention it is in Python?
used by the ID division at St.\ Michael's Hospital~\cite{landsman_scheduling}. 
The software was used to
generate the results in the following sections, using real data as well as
simulated data as input. All the following experiments were conducted on an
Intel Core i7-4770k CPU @ 3.50 GHz with 16 GB of RAM running 64-bit Windows 10.
Our software package uses COIN-OR Branch-and-Cut open source solver
% JK: "software" here is not clear to me, is it the solver for the ILP formulation?
version 2.9.9~\cite{johnjforrest_coin-or/cbc:_2019}.

\subsection{Comparison with Manually Generated Schedules}  %main comment  = I think use past
%tense in writing
We used clinician time-off requests and minimum/maximum requirements from
2015-2018 as input data for the ILP problem.
Table~\ref{tbl:2018-schedule-comparison} compares the optimal schedule generated using
the software with the manually-created schedule for data from 2018. The
schedule is color-coded to distinguish between the different clinicians.

\input{tbl/2018_schedule_comparison}
% JK: I think use input not include here to avoid unnecessary page breaks
%     https://tex.stackexchange.com/questions/246

First, we evaluated the ILP solution by comparing it with the
% JK: I think it would be good to give a short name for manual and the proposed
%     data -- e.g. LP Solution as in Table 4 and Manual Generation.
%     Then check for consistency throughout.
%     While lack of synonyms can sound repetitive, it can help avoid confusion.
manual generation, as in Table~\ref{tbl:2018-schedule-comparison}. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the ILP solution satisfied all 
hard constraints. In contrast,
% JK: "mandatory" = "hard"? if possible just use one.
manual generation did not satisfy all hard constraints. In
particular, we see that the manual generation assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual generation did not have
an equal distribution of weekends and holidays for all four years of data.
Considering all objectives, we see that the ILP solution outperforms manual generation
in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks.

\input{tbl/constraints_comparison}
% JK: If you want to be funny, you can estimate the approximate time to
%     generate the schedule and include that too (16 hours vs 1 second)...

@@

% JK: Introduce what you did first and then the figure showing results (see above)
The effect of increasing the number of clinicians and number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}.
We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. 
In a department providing a single service, increasing the number of clinicians
did not affect the runtime, and we were able to find an ILP solution in all four cases
within 1 second.
For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when removing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds.  %what does division mean? do you mean services?

@@

% JK: Good to introduce the detault parameters first, then for each parameter explored,
%     Describe what varied.
For the remaining experiments, we simulated a department with 10 clinicians offering 
two services, similar to the department at St.\ Michael's Hospital. 
% JK: \abs{BR} does not clearly imply number of blocks (could be confused with week indices),
%     so just leave out the LHS here I think.
% JK: What do you mean by non-overlapping? And why only increments of 5?
The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}.
In this experiment, each clinician was configured with 1 to 15 total block requests.
The runtime of the algorithm is constant with respect to the number of requests,
indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all runs were completed in under
2 seconds.
% JK: See somments below about trends and repeating simulations.

\begin{figure}[h]
	\centering
	\def\svgwidth{\columnwidth}
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
	\input{fig/runtime-requests.pdf_tex} % JK: can you export this to PDF or SVG to avoid pixelation?
  \label{fig:runtime-requests}
  % JK: The trend here is a bit confusing, since I think you're trying to say
  %     that the runtime is not affected by the number of constraints,
  %     but there is noise in these results.
  %     I would re-run this 10-100 times and average the results so that (I expect)
  %     the mean trend is actually flat. You could (should) then also present the runtimes
  %     as box-plots to show that the trend is less than the variability.
\end{figure}

% JK: Introduce what you did first and then the figure showing results (see above)
% JK: Try to introduce these results in the same order as they appear in the
%     first paragraph of this section.
Figure~\ref{fig:runtime-blocks} presents the change in runtime when increasing
the number of 2-week blocks in a department with 10
clinicians offering two services. In this experiment, we investigated time horizons
from 5 to 110 blocks. This is equivalent to generating a schedule for up to 4 years ahead.
The trend in the graph indicates a linear growth in runtime with respect to the time-horizon.
Notably, the ILP solver was able to find all solutions in under 6 seconds, indicating very 
good performance for long-term scheduling. 
% 'very reasonable amount of time" is vague. what is reasonable?
%better to give a range in value.
% JK: same comment as above re. trends

\begin{figure}[h]
	\centering
	\def\svgwidth{\textwidth}
	\caption{Runtime of ILP solver with an increasing number of 2-week blocks}%
	\input{fig/runtime-blocks.pdf_tex}
  \label{fig:runtime-blocks}
  % JK: same comment as above re. trends
  % JK: Thess figures could also possible be Figure 1. a) and b) 
\end{figure}

>>>>

Our goal was to determine if the schedules provided by solving the ILP 
could successfully
(i) enforce all hard constraints; 
(ii) improve fulfillment of soft constraints compared to the manual approach; 
and (iii) assess whether our ILP formulation can be used for a wide range
of configurations.
First, we compared the schedules created by solving the ILP formulation
given in Section~\ref{sec:methods} to schedules that were manually generated,
with respect to adherence to the hard and soft constraints outlined in Section~\ref{sec:problem}.
We then examined the efficiency of the ILP approach in generating schedules
by its runtime on a variety of instances that may be found in the real-world.

\subsection{Implementation}
We developed a Python software package with a user interface that implements the above
linear program and allows configuration of clinicians, to be
used by the ID division at St.\ Michael's Hospital~\cite{landsman_scheduling}. 
The software was used to
generate the results in the following sections, using real data as well as
simulated data as input. All the following experiments were conducted on an
Intel Core i7-4770k CPU @ 3.50 GHz with 16 GB of RAM running 64-bit Windows 10.
Our software package uses COIN-OR Branch-and-Cut open source solver
version 2.9.9~\cite{johnjforrest_coin-or/cbc:_2019}.

\subsection{Comparison with Manually Generated Schedules}
We used clinician time-off requests and minimum/maximum requirements from
2015-2018 as input data for the ILP problem.
Table~\ref{tbl:2018-schedule-comparison} compares the optimal schedule generated using
the software with the manually-created schedule for data from 2018. The
schedule is color-coded to distinguish between the different clinicians.

\input{tbl/2018_schedule_comparison}

First, we evaluated the ILP solution by comparing it with the
manual generation, as in Table~\ref{tbl:2018-schedule-comparison}. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the ILP solution satisfied all 
hard constraints. In contrast,
manual generation did not satisfy all hard constraints. In
particular, we see that the manual generation assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual generation did not have
an equal distribution of weekends and holidays for all four years of data.
Considering all objectives, we see that the ILP solution outperforms manual generation
in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks.

\input{tbl/constraints_comparison}

@@

The effect of increasing the number of clinicians and number of services 
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}.
We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in  
total across all services. 
In a department providing a single service, increasing the number of clinicians
did not affect the runtime, and we were able to find an ILP solution in all four cases
within 1 second.
For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when removing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds.

@@

For the remaining experiments, we simulated a department with 10 clinicians offering 
two services, similar to the department at St.\ Michael's Hospital. 
The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}.
In this experiment, each clinician was configured with 1 to 15 total block requests.
The runtime of the algorithm is constant with respect to the number of requests,
indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all runs were completed in under
2 seconds.

\begin{figure}[h]
	\centering
	\def\svgwidth{\columnwidth}
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
	\input{fig/runtime-requests.pdf_tex}
  \label{fig:runtime-requests}
\end{figure}

Figure~\ref{fig:runtime-blocks} presents the change in runtime when increasing
the number of 2-week blocks in a department with 10
clinicians offering two services. In this experiment, we investigated time horizons
from 5 to 110 blocks. This is equivalent to generating a schedule for up to 4 years ahead.
The trend in the graph indicates a linear growth in runtime with respect to the time-horizon.
Notably, the ILP solver was able to find all solutions in under 6 seconds, indicating very 
good performance for long-term scheduling.

\begin{figure}[h]
	\centering
	\def\svgwidth{\textwidth}
	\caption{Runtime of ILP solver with an increasing number of 2-week blocks}%
	\input{fig/runtime-blocks.pdf_tex}
  \label{fig:runtime-blocks}
\end{figure}

### 2019-10-11 paper/paper.intro.tex
Cleanup latex comments

Hospital departments must allocate and use their limited resources efficiently
in order to provide a high quality of care for their patients.
%Hospital departments provide services where patient needs, and thus the system's
%demands, often exceed the available supply. 
% JK: Not sure I follow this sentence.
%     I think we dont need to say demands > needs, can just say efficieny improves care.
In particular, on-call schedules for
a fixed number of health-care providers are central to the efficient running of
hospitals. Carefully allocated schedules should balance
sufficient staff with clinician workload,
to maximize quality of care. It is common for on-call schedules in
hospitals to be created manually. However,
manually-created schedules are subject to three problems.
First, when there are a large number of clinicians,
or the constraints that need to be satisfied
by the schedule are complex, it becomes infeasible
to find a satisfying schedule by hand.
%a manual method may not provide a schedule
%that satisfies all constraints. 
Second, when creating a schedule manually it is difficult to ensure
that all constraints are met while also trying to satisfy all staff
preferences.
%manual methods may overlook
%certain constraints that must be maintained to have an operational department,
%such as preventing many consecutive work blocks from being assigned or ensuring
%clinicians are allocated a specific amount of work blocks throughout the year,
%especially if there are many complex constraints to keep track of.
% JK: Can you clarify why / how these may be overlooked?
Third, manual scheduling is often time-consuming even for relatively small 
departments, and can take up time and resources that are better used
for improving patient care.
For these reasons, it is important to develop automated methods that
can efficiently generate schedules that satisfy the given constraints.
% JK: haven't introduced what "optimal" means yet so I think maybe leave it out for now.
%For example, it is important that a hospital department allocates its
%resources, such as the availability of a finite number of clinicians,
%optimally, to ensure the best possible service for its patients.
% JK: instead of using \\ everywhere for paragraph breaks, use:
%     \setlength{\parskip}{1em}
%     in the preamble

Automated methods to generate schedules have been studied and applied in many
industries, including
% JK: I find this is a nice way to organize these lists of topics & citations
transportation~\cite{aickelin_improved_2006, goel_truck_2012, gunther_combined_2010},
manufacturing~\cite{al-yakoob_mixed-integer_2007, al-yakoob_column_2008, alfares_simulation_2007},
retail~\cite{chapados_retail_2011, nissen_automatic_2010}, and
military~\cite{horn_scheduling_2007, laguna_modeling_2005}.
Of special interest to a clinician scheduling
problem are the approaches to scheduling nurses, who often work in shifts. In the
nurse scheduling problem, the goal is to find an optimal assignment of nurses to
shifts that satisfies all hard constraints (such as hospital regulations),
and as many soft constraints (such as nurse preferences) as possible.
% JK: Is this specific to the nurse scheduling problem?
%     This sounds like the more general definition of the scheduling problem.
%     Can you define hard and soft constraints here and give
%     one example of a hard constraint and one example of a soft constraint.
Hard constraints must be satisfied by any candidate solution to the nurse scheduling problem,
while soft constraints can be used to rank the candidate solutions. 
For instance, a nurse scheduling problem may include a hard constraint to assign
at most a single shift for each nurse per day. It can also incorporate 
nurse preference for shift time (that is, day versus night shifts) as a soft constraint
that is meant to optimize the schedule, but is not guaranteed to be fulfilled.
A wide variety of approaches, including exact and heuristic approaches, have been
used to solve the nurse scheduling problem:
integer linear programming~\cite{azaiez_0-1_2005, trilling_nurse_2006, widyastiti_nurses_2016},
network flows~\cite{el_adoly_new_2018},
genetic algorithms~\cite{aickelin_exploiting_2000, jan_evolutionary_2000, kawanaka_genetic_2001},
simulated annealing~\cite{jaszkiewicz_metaheuristic_1997}, and
artificial intelligence~\cite{abdennadher_nurse_1999, li_hybrid_2003}.
A comprehensive
literature review of these and other methods applied to nurse scheduling is
presented in~\cite{burke_state_2004}.
% JK: rostering = scheduling? I think okay to re-use "scheduling" unless its getting extremely repetitive.
%     but even then I think SM would agree clarity > flow/repetition.

%An extensive literature review of these and other methods is presented by [??].
%We will briefly summarize the main ideas of some of these approaches. \\ SM -
%don't need to introduce that you will do this for this type of paper I think -
%would do in a thesis chapter though.

Many of the approaches to nurse scheduling were designed to satisfy the
requirements of a specific hospital department, which results in a large number of
variables and constraints to be incorporated into the problem formulation. While
these department-specific approaches allow end-users to find precise schedules
that satisfy the needs of that department and the preferences of nurses and
clinicians in that department, they are difficult to adapt to other
departments. % explain why hard to
%adapt?... what makes their generalizablility/adaptability limited?
% JK: I'm less bothered by this lack of explaination, since I know its
%     hard to explain without getting into the nitty-gritty.
Moreover, the large number of variables and constraints also leads to
computational complexity issues~\cite{goos_complexity_1996}, especially when
% JK: I think good to clarify that these complextiy issues would imply that
%     the problem is essentially impossible to solve in a limited amount of time.
trying to find the most optimal solution. In particular, difficult instances of these formulations
become impossible to solve in a reasonable amount of time. 
In this paper, we tackle the clinician 
scheduling problem arising from a case study of one clinical
division, providing two different services (general infectious
disease (ID) consults; and HIV consults) at St.\ Michael's Hospital in
Toronto, Canada. The clinician scheduling problem involves creating
a yearly schedule that assigns clinicians to on-call work on a weekly basis,
while ensuring a fair and balanced workload.
% JK: the nurse scheduling problem?
%     I think we need to define the "scheduling problem"
%     and how / if it differs from other scheduling problem.
Our goals in this paper are to (1) present an integer linear programming 
(ILP) formulation for our problem, and
describe the flexibility of this formulation for solving similar problems;
(2) compare the performance of our tool for solving the ILP formulation to the
results of a manual approach; 
and (3) analyze the robustness of this approach
in difficult instances of the problem, by exploring the change in runtime with
changes to:
the number of clinicians,
the number of services provided by the department,
the number of requests per clinician, and
the time-horizon of the schedule.
% JK: what do you mean by different instances?  Can you be more specific?
%     The results explore:
%     (1) runtime vs number of requests
%     (2) runtime vs number of blocks
%     (3) satisfaction of hard and soft constraints.
% present a simple formulation for the problem developed and tested at the
%hospital after switching from a manual approach to scheduling; and (2) analyze
%the performance of integer linear programming in solving difficult instances of
%the problem and compare the results with those of the manual approach; and (3)
%describe the adaptability of the formulation as a basic framework for solving
%similar problems in other departments. \\ %make #3 an objective

We begin by describing the details of the problem in Section~\ref{sec:problem}, and presenting
our ILP formulation in Section~\ref{sec:methods}. Next, we compare the results
of our formulation to manually-created schedules, and evaluate the performance
of the algorithm on simulated data in Section~\ref{sec:experiments}. Finally, we
discuss and interpret the results in Section~\ref{sec:discussion}. % list the
%main contents/elements of the paper.

>>>>

Hospital departments must allocate and use their limited resources efficiently
in order to provide a high quality of care for their patients.
In particular, on-call schedules for
a fixed number of health-care providers are central to the efficient running of
hospitals. Carefully allocated schedules should balance
sufficient staff with workload
to maximize quality of care. It is common for on-call schedules in
hospitals to be created manually. However,
manually-created schedules are subject to three problems.
First, when there are a large number of clinicians,
or the constraints that need to be satisfied
by the schedule are complex, it becomes infeasible
to find a satisfying schedule by hand. 
Second, when creating a schedule manually it is difficult to ensure
that all constraints are met while also trying to satisfy all staff
preferences.
Third, manual scheduling is often time-consuming even for relatively small 
departments, and can take up time and resources that are better used
for improving patient care.
For these reasons, it is important to develop automated methods that
can efficiently generate schedules that satisfy the given constraints.

Automated methods to generate schedules have been studied and applied in many
industries, including
transportation~\cite{aickelin_improved_2006, goel_truck_2012, gunther_combined_2010},
manufacturing~\cite{al-yakoob_mixed-integer_2007, al-yakoob_column_2008, alfares_simulation_2007},
retail~\cite{chapados_retail_2011, nissen_automatic_2010}, and
military~\cite{horn_scheduling_2007, laguna_modeling_2005}.
Of special interest to a clinician scheduling
problem are the approaches to scheduling nurses, who often work in shifts. In the
nurse scheduling problem, the goal is to find an optimal assignment of nurses to
shifts that satisfies all hard constraints (such as hospital regulations),
and as many soft constraints (such as nurse preferences) as possible.
Hard constraints must be satisfied by any candidate solution to the nurse scheduling problem,
while soft constraints can be used to rank the candidate solutions. 
For instance, a nurse scheduling problem may include a hard constraint to assign
at most a single shift for each nurse per day. It can also incorporate 
nurse preference for shift time (that is, day versus night shifts) as a soft constraint
that is meant to optimize the schedule, but is not guaranteed to be fulfilled.
A wide variety of approaches, including exact and heuristic approaches, have been
used to solve the nurse scheduling problem:
integer linear programming~\cite{azaiez_0-1_2005, trilling_nurse_2006, widyastiti_nurses_2016},
network flows~\cite{el_adoly_new_2018},
genetic algorithms~\cite{aickelin_exploiting_2000, jan_evolutionary_2000, kawanaka_genetic_2001},
simulated annealing~\cite{jaszkiewicz_metaheuristic_1997}, and
artificial intelligence~\cite{abdennadher_nurse_1999, li_hybrid_2003}.
A comprehensive
literature review of these and other methods applied to nurse scheduling is
presented in~\cite{burke_state_2004}.

Many of the approaches to nurse scheduling were designed to satisfy the
requirements of a specific hospital department, which results in a large number of
variables and constraints to be incorporated into the problem formulation. While
these department-specific approaches allow end-users to find precise schedules
that satisfy the needs of that department and the preferences of nurses and
clinicians in that department, they are difficult to adapt to other
departments.
Moreover, the large number of variables and constraints also leads to
computational complexity issues~\cite{goos_complexity_1996}, especially when
trying to find the most optimal solution. In particular, difficult instances of these formulations
become impossible to solve in a reasonable amount of time. 
In this paper, we tackle the clinician 
scheduling problem arising from a case study of one clinical
division, providing two different services (general infectious
disease (ID) consults; and HIV consults) at St.\ Michael's Hospital in
Toronto, Canada. The clinician scheduling problem involves creating
a yearly schedule that assigns clinicians to on-call work on a weekly basis,
while ensuring a fair and balanced workload.
Our goals in this paper are to (1) present an integer linear programming 
(ILP) formulation for our problem, and
describe the flexibility of this formulation for solving similar problems;
(2) compare the performance of our tool for solving the ILP formulation to the
results of a manual approach; 
and (3) analyze the robustness of this approach
in difficult instances of the problem, by exploring the change in runtime with
changes to:
the number of clinicians,
the number of services provided by the department,
the number of requests per clinician, and
the time-horizon of the schedule.

We begin by describing the details of the problem in Section~\ref{sec:problem}, and presenting
our ILP formulation in Section~\ref{sec:methods}. Next, we compare the results
of our formulation to manually-created schedules, and evaluate the performance
of the algorithm on simulated data in Section~\ref{sec:experiments}. Finally, we
discuss and interpret the results in Section~\ref{sec:discussion}.
